Interactive Visual Analysis of Image-Centric Cohort Study Data

### 2014-03-21 draft.tex
first part of the Application section

%% Abstract section.
\abstract{%
%(The problem). 
Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease-specific risk factors.
%
Cohort studies comprise heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors.
%
The data is analyzed using a priori defined hypotheses to find statistically significant correlations between variables (\emph{interactions}).
%
Modern cohort studies also incorporate medical image data.
%
%Analyzing these data requires image segmentation, extraction of key figures, and shape based subject grouping.
%
The statistically driven epidemiological workflow only allows to determine \emph{interactions} between image-derived metrics such as distances extracted from landmarks of the segmentation model.
\\\\
%(New Solution).
We propose an Interactive Visual Analysis approach that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic variables and attributes derived from the image data.
%
This is achieved by combining brushing and linking enabled coordinated information visualization views and interactive 3D shape renderings with epidemiological data representations such as pivot tables and key figures as odds ratios.
%incorporating coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
%
%Multiple coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
%
%(Validation).
The presented concepts are applied expert-guided to gather and evaluate hypotheses about the aging process of the lumbar spine.
%
%(Results). 
It shows to be a more flexible comparison between image and non-image data.
%
%(Implications).
The new framework allows for hypothesis validation and hypothesis generation by incorporating human pattern recognition as well as data mining methods.
%
Using all reliable information from the image segmentation linked to non-image variables aims to unveil \emph{interactions} by applying an iterative analysis approach.

@@

\section{Interactive Visual Analytics in Cohort Study Data}
\begin{figure}[htb]
 \centering
 \label{fig:WorkflowComparison}
 \includegraphics[width=3.0in]{figures/workflow_comparison}
 \caption{ToDo Abbildung noch nicht final. Visual Analytics systems are able to complement parts of the epidemiological workflow, not replace it. The appropriate combination of statistical- and interactive driven analysis shows promising potential to unveil the information in the data. (a) shows the standard epidemiological workflow, (b) the IVA supported one. The iterative red highlighted part is called the \emph{IVA Loop}.}
\end{figure}
%
As described in subsection~\ref{EpidemiologicalWorkflow}, the epidemiolgical workflow is a sequence of steps taken by domain experts and needs to be reproducible and comprise statistical integrity.
%
Figure~\ref{fig:WorkflowComparison} (a) describes this workflow as consecutive series of steps.
%
The workflow we propose by introducing the \emph{IVA} principle into the epidemiological application domain does not aim to replace the existing workflow but to complement its weaknesses.
%
In the current state the workflow treats the data like a black box.
%
A list of features describing the hypothesis is compiled and analyzed using statistical tests. 
%
The resulting value decides whether the data supports the hypothesis or not.
%
It would be possible that there are actually features of the data set which support the hypothesis by discriminating the population in the expected way, but with this approach they are not highlighted in any way.
%
This becomes even more important when the workflow is adapted to the analysis of the medical image data.
%
Domain experts would have to annotate landmarks tediously to derive metrics such as distances. %which are then handled like other features and analyzed using the same set of statistical tools.
%
Not only does this leave out the majority of the information in the medical image data by abstracting it to single values, it is easily possible that information left out would heavily influence the result.
%
Considering more complex parts of the data would make those results more trustworthy and also could identify possible anatomical confounders--an epidemiolgical research result in itself.
%
Statistical tests check for validity of the number but not for their completeness or plausibility!
\\\\
\emph{IVA} tries to illuminate the black box by making the domain experts part of the feature list selection process.
%
Figure~\ref{fig:WorkflowComparison} (b) highlights the iterative process as part of the epidemiolgical workflow.
%
Note that it also aims to project back into the hypothesis formulation step to amplify hypothesis generation 
%
This has to be handled with care since overfitting of expectations to the available data is an imminent danger as described by Turkay and colleagues \cite{Turkay2013}.

@@

\section{Application}
- Acceptance plays a big role - you have to pick up the user by presenting information they are used to in a new way
\subsection{The Spine Dataset}
- Describe steps from gathering Information from the raw image files (segmentation, abstraction, visualization)\\
- Problem of sparse differences - Visualization has to be more abstract to emphasize the differences.\\
- Input of Epidemiologists goes here!\\
\\\\
\textbf{From VMV'13 Paper}\\

All whole-body MRI scans were acquired on a 1.5~Tesla scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany) by four trained technicians in a standardized way. Subjects were placed in the supine position. Five phased-array surface coils were placed to the head, neck, abdomen, pelvis, and lower extremities for whole-body imaging. The spine coil is embedded in the patient table. The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence (676 / 12 [repetition time msec / echo time msec]; 150$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence (3760 / 106 [repetition time msec / echo time msec]; 180$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels). First, both sequences were placed over the cervical and upper thoracic spine. Then, they were placed over the lower thoracic and lumbar spine. The MRI software automatically composed a whole spine sequence from the two T1-weighted and T2-weighted sequences \cite{Hegenscheid2013}. We were provided with 490 data sets.

%The model is placed in the scene using an empirically chosen initialization point. The force acting on the model stems from aggregation of loads, which are derived from a potential field resulting from a weighted sum of the T1- and T2-weighted MRI images, see \cite{Rak2013}. After detecting all spines, we register the models because in a later clustering step we only want to capture the local deformation of the lumbar spine, not different locations in world space. The models are registered using the Kabsch Algorithm \cite{Kabsch1976}, which is designed to minimize the root mean squared deviation between paired sets of points.
The model is placed in the scene using an empirically chosen initialization point. The force acting on the model stems from aggregation of loads, which are derived from a potential field resulting from a weighted sum of the T1- and T2-weighted MRI images, see \cite{Rak2013}. After detecting all spines, we register the models because in a later clustering step we only want to capture the local deformation of the lumbar spine, not different locations in world space. The models are registered using the Kabsch Algorithm, which is designed to minimize the root mean squared deviation between paired sets of points.
The model-based detection captures information about the spine canal curvature as well as the alignment of the vertebrae. It is not meant to capture information about vertebrae deformation and differences in spine canal extent.

>>>>

%% Abstract section.
\abstract{%
%(The problem). 
Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease-specific risk factors.
%
Cohort studies comprise heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors.
%
The data is analyzed using a priori defined hypotheses to find statistically significant correlations between variables (\emph{interactions}).
%
Modern cohort studies also incorporate medical image data.
%
%Analyzing these data requires image segmentation, extraction of key figures, and shape based subject grouping.
%
The statistically driven epidemiological workflow only allows to determine \emph{interactions} between image-derived metrics such as distances extracted from landmarks of the segmentation model.
\\\\
%(New Solution).
We propose an Interactive Visual Analysis approach that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic variables and attributes derived from the image data.
%
This is achieved by combining brushing and linking enabled coordinated information visualization views and interactive 3D shape renderings with epidemiological data representations such as pivot tables and key figures as association measures.
%incorporating coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
%
%Multiple coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
%
%(Validation).
The presented concepts are applied expert-guided to gather and evaluate hypotheses about the aging process of the lumbar spine.
%
%(Results). 
It shows to be a more flexible comparison between image and non-image data.
%
%(Implications).
The new framework allows for hypothesis validation and hypothesis generation by incorporating human pattern recognition as well as data mining methods.
%
Using all reliable information from the image segmentation linked to non-image variables aims to unveil \emph{interactions} by applying an iterative analysis approach.

@@

\section{Interactive Visual Analyis in Cohort Study Data}
\begin{figure}[htb]
 \centering
 \label{fig:WorkflowComparison}
 \includegraphics[width=3.0in]{figures/workflow_comparison}
 \caption{ToDo Abbildung noch nicht final. Visual Analytics systems are able to complement parts of the epidemiological workflow, not replace it. The appropriate combination of statistical- and interactive driven analysis shows promising potential to unveil the information in the data. (a) shows the standard epidemiological workflow, (b) the IVA supported one. The iterative red highlighted part is called the \emph{IVA Loop}.}
\end{figure}
- Acceptance plays a big role - you have to pick up the user by presenting information they are used to in a new way\\
%
As described in subsection~\ref{EpidemiologicalWorkflow}, the epidemiolgical workflow is a sequence of steps taken by domain experts and needs to be reproducible and comprise statistical integrity.
%
Figure~\ref{fig:WorkflowComparison} (a) describes this workflow as consecutive series of steps.
%
The workflow we propose by introducing the \emph{IVA} principle into the epidemiological application domain does not aim to replace the existing workflow but to complement its weaknesses.
%
In the current state the workflow treats the data like a black box.
%
A list of features describing the hypothesis is compiled and analyzed using statistical tests. 
%
The resulting value decides whether the data supports the hypothesis or not.
%
It would be possible that there are actually features of the data set which support the hypothesis by discriminating the population in the expected way, but with this approach they are not highlighted in any way.
%
This becomes even more important when the workflow is adapted to the analysis of the medical image data.
%
Domain experts would have to annotate landmarks tediously to derive metrics such as distances. %which are then handled like other features and analyzed using the same set of statistical tools.
%
Not only does this leave out the majority of the information in the medical image data by abstracting it to single values, it is easily possible that information left out would heavily influence the result.
%
Considering more complex parts of the data would make those results more trustworthy and also could identify possible anatomical confounders--an epidemiolgical research result in itself.
%
Statistical tests check for validity of the number but not for their completeness or plausibility!
\\\\
\emph{IVA} tries to illuminate the black box by making the domain experts part of the feature list selection process.
%
Figure~\ref{fig:WorkflowComparison} (b) highlights the iterative process as part of the epidemiolgical workflow.
%
Note that it also aims to project back into the hypothesis formulation step to amplify hypothesis generation 
%
This has to be handled with care since overfitting of expectations to the available data is an imminent danger as described by Turkay and colleagues \cite{Turkay2013}.

@@

\section{Application}
We apply the presented set of techniques to a data set which is compiled to analyze lower back pain which is one of the most common reasons for an adult to see a physicians \cite{Backpain}.
%
Data confidentiality and ethical reasons prohibit us from accessing the complete data space of the SHIP feature space.
%
The epidemiologists compiled feature list which is a tradeoff between complexity and ability to pass the ethics committee.
%
%The goal is to find features which \emph{interact} with lower back pain based on image and non-image data.
%
\subsection{The Spine Dataset}
We divide the data set in image and non-image data.
%
There are 136 features describing diagnosed diseases, lifestyle factors, women specific factors, pain indicators, laboratory values somatometric variables and are ordered accordingly.
%
The image data was acquired on a 1.5~Tesla scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany) by four trained technicians in a standardized way.
%
The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence (676 / 12 [repetition time msec / echo time msec]; 150$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence (3760 / 106 [repetition time msec / echo time msec]; 180$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.

\subsection{Processing the Image data}
The lumbar spine was detected in the image data using a hierarchical finite element method according to \cite{Rak2013}.
%
This semi-automatic method requires the user to initialize the Tetrahedron-based finite element models (FEM) with a click on the L2-vertebra.
%
It uses a weighted sum of T1- and T2-weighted MRI images to detect the lumbar spine shape.
%
The registered models capture resilient information about shape of the lumbar spine canal as well as the position of the L1 -L5 vertebrae.
%
For clustering purposes we extracted the centerline of the spine canal of the lumbar spine canal which captures information about lordosis and scoleoses which is the medical term for spine curvature \cite{Klemm2013VMV}.
%
%This information is a major feature for diagnosing lumbar back pain.
\subsection{Processing the Non-Image data}
- SPSS Export? JSON-Export? Automatisches Anfuegen des Dictionaries?

%\subsection{The Spine Prototype}

- Clustering using Steffens Algorithm
% - Describe steps from gathering Information from the raw image files (segmentation, abstraction, visualization)\\
% - Problem of sparse differences - Visualization has to be more abstract to emphasize the differences.\\
% - Input of Epidemiologists goes here!\\
% \\\\
% \textbf{From VMV'13 Paper}\\
% 
% All whole-body MRI scans were acquired on a 1.5~Tesla scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany) by four trained technicians in a standardized way. Subjects were placed in the supine position. Five phased-array surface coils were placed to the head, neck, abdomen, pelvis, and lower extremities for whole-body imaging. The spine coil is embedded in the patient table. The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence (676 / 12 [repetition time msec / echo time msec]; 150$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence (3760 / 106 [repetition time msec / echo time msec]; 180$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels). First, both sequences were placed over the cervical and upper thoracic spine. Then, they were placed over the lower thoracic and lumbar spine. The MRI software automatically composed a whole spine sequence from the two T1-weighted and T2-weighted sequences \cite{Hegenscheid2013}. We were provided with 490 data sets.

% The model is placed in the scene using an empirically chosen initialization point. The force acting on the model stems from aggregation of loads, which are derived from a potential field resulting from a weighted sum of the T1- and T2-weighted MRI images, see \cite{Rak2013}. After detecting all spines, we register the models because in a later clustering step we only want to capture the local deformation of the lumbar spine, not different locations in world space. The models are registered using the Kabsch Algorithm, which is designed to minimize the root mean squared deviation between paired sets of points.
% The model-based detection captures information about the spine canal curvature as well as the alignment of the vertebrae. It is not meant to capture information about vertebrae deformation and differences in spine canal extent.
